---
title: "1. RNAseq analysis of FACS-purified ISC/EBs"
description: "DGE analysis based on DESeq2"
principal investigator: "Joaquín de Navascués"
researchers: "Aleix Puig-Barbé, Joaquín de Navascués"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: readable
    df_print: paged
    css: doc.css
---
```{r setup, echo = FALSE, cache = FALSE}
ggplot2::theme_set(ggpubr::theme_pubr(base_size=10))
fsep <- .Platform$file.sep
knitr::opts_chunk$set(dev = 'png', 
                      fig.align = 'center', fig.height = 7, fig.width = 10, 
                      pdf.options(encoding = "ISOLatin9.enc"),
                      fig.path=paste0('notebook_figs', fsep), warning=FALSE, message=FALSE)
```


# 1 Preparation


**Libraries:**
```{r libraries, warning=FALSE, message=FALSE}
if (!require("librarian")) install.packages("librarian")
librarian::shelf(dplyr, stringr,
                 DESeq2, edgeR, biomaRt, rtracklayer, GenomicFeatures, limma,
                 ggplot2, pheatmap, scico, dendsort,
                 here, writexl, gzcon,
                 quiet = TRUE)
```

**Set working directory:**
```{r setwd}
fsep <- .Platform$file.sep
if (Sys.getenv("RSTUDIO")==1) {
   # setwd to where the editor is, if the IDE is RStudio
  setwd( dirname(rstudioapi::getSourceEditorContext(id = NULL)$path) )
} else {
  # setwd to where the editor is in a general way - maybe less failsafe than the previous
  setwd(here::here())
  # the following checks that the latter went well, but assuming
  # that the user has not changed the name of the repo
  d <- str_split(getwd(), fsep)[[1]][length(str_split(getwd(), fsep)[[1]])]
  if (d != 'Puigetal2023_bioinformatics_scripts') { stop(
    paste0("Could not set working directory automatically to where this",
           " script resides.\nPlease do `setwd()` manually"))
    }
}
```

**To save images outside the repo (to reduce size):**
```{r define_dir2figs}
figdir <- paste0(c(head(str_split(getwd(), fsep)[[1]],-1),
                   paste0(tail(str_split(getwd(), fsep)[[1]],1), '_figures')),
                 collapse = fsep)
dir.create(figdir, showWarnings = FALSE)
```


## 1.1 Experimental conditions


These are defined in detail in the paper as well as the GEO submission.
In an nutshell, five different strains were crossed to the _esg-Gal4, UAS-GFP_ driver stock, in two batches:

* Batch `1`:
  * control
  * _UAS-daughterless_ (_da_)
  * _UAS-da^RNAi^_ (TRiP-line, either <span style="color: red;">**JF02488 or JF02092**</span>)
* Batch `2`:
  * control
  * _UAS-da:da_
  * _UAS-scute_

The target cells were FAC-sorted, their RNA extracted, reverse-transcribed, amplified and sequenced with Illumina as described in [Korzelius et al. (2019)](https://www.nature.com/articles/s41467-019-12003-0).


## 1.2 Get gene symbols


Genes are now identified as FlyBase IDs (e.g. FBgn0000061). To get also the gene names (e.g. _aristaless_):
```{r get-gene-symbols}
ensembl <- useEnsembl(biomart = "ENSEMBL_MART_ENSEMBL",
                      dataset="dmelanogaster_gene_ensembl",
                      host = "https://oct2022.archive.ensembl.org")
                      # to update this: https://www.ensembl.org/Help/ArchiveRedirect
filters <- listFilters(ensembl) # define filters for a specific query
attributes <- listAttributes(ensembl) # define the features showed

dlist <- getBM(attributes = c('ensembl_gene_id', 'external_gene_name'),
               mart = ensembl)
rownames(dlist) <- dlist$ensembl_gene_id
dlist[1] <- NULL
names(dlist) <- 'gene_symbol'
write.table(dlist, file=file.path(getwd(), "resources", "gene_symbols.txt"), col.names=NA)
```


## 1.3 Load raw count data


### 1.3.1 Download the pre-processed data

The datasets are available in GEO. We can simply get a list of download URLs like this:
```{r get-counts-geo}
# get experimental design data:
inpath <- file.path(getwd(), 'input')
samples <- list('C3N4AACXX_1', 'C3N4AACXX_4', 'C3N4AACXX_9', 'C3N4AACXX_7',
                'C3N4AACXX_11', 'C3N4AACXX_2', 'C3N4AACXX_6', 'C3N4AACXX_10',
                'CON1', 'CON2', 'CON3', 'DA1', 'DA2', 'DA3', 'SCUTE1',
                'SCUTE2', 'SCUTE3')
ids <- paste0('GSM7441', 184:200)
targets <- read.table(file.path(inpath, "targets.txt"), header=TRUE, sep="\t")
targets$GEOsample <- ids[match(targets$sampleID, samples)]
# download data or read cache
if (file.exists(file.path(inpath, 'rawData.rds'))) {
  rawData <- readRDS(file.path(inpath, 'rawData.rds'))
} else {
  # prepare download URLs
  fore <- 'https://www.ncbi.nlm.nih.gov/geo/download/?acc='
  mid <- '&format=file&file='
  hind <- '.featurecount.txt.gz'
  urls <- paste0(fore, ids, mid, ids, '_', samples, hind)
  names(urls) <- ids
  # prepare to load gene count data
  rawData <- NULL
  # each column of rawData will contain the reads per gene of a sample
  counter <- 0
  for (geo in targets$GEOsample) {
    connlink <- url( urls[ names(urls)==geo ] )
    on.exit(close(connlink))
    conn <- gzcon(connlink)
    txt <- readLines(conn)
    fileContents <- read.table(textConnection(txt), sep="\t", header=T)
    rawData <- cbind(rawData, fileContents[,7])
  }
  # add column and row names to the `rawData` matrix
  colnames(rawData) <- paste(targets$Condition, targets$Replicate, targets$Batch, sep='_')
  rownames(rawData) <- fileContents$Geneid
  # remove genes with low counts
  cpms <- cpm(rawData)
  keep <- rowSums(cpms > 1) >= 3 # detected in at least 3 samples
  rawData <- rawData[keep,]
  # cache
  saveRDS(rawData, file.path(inpath, 'rawData.rds'))
}
```


# 2 Differential gene expression


## 2.1 Set up `DESeq2DataSet` object


Create experimental design object that contains the information from `targets`:
```{r expt-design}
exptDesign = data.frame(
  row.names = colnames(rawData),
  condition = targets$Condition,
  batch = targets$Batch)
```

Create `DESeq2DataSet` object with information from `exptDesign` and `rawData`:
```{r dge, warning=FALSE}
exptObject <- DESeqDataSetFromMatrix(countData = rawData,
                                     colData = exptDesign,
                                     design = ~ batch + condition) 
# specify 'Control' as the reference level
exptObject$condition <- relevel(exptObject$condition, ref = "Control")
```
The formula `~ batch + condition` allows to take into account the genotype of the samples as well as the batch they come from, to remove batch effects. Note that `condition` is last in the formula - this is needed for the genotype to be the predictor variable (see [here](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#differential-expression-analysis) why).


## 2.2 Batch-correction


Transform the normalized counts and plot them as PCA:
```{r pca-raw}
vsd_Object <- vst(exptObject, blind=TRUE)
plotPCA(vsd_Object)
```

There's a clear (and expected) batch effect: batch 1 is in PC1 negative values while batch 2 is in PCA1 positive values. Remove batch effect and re-plot:
```{r pca-batch-corr}
assay(vsd_Object) <- removeBatchEffect(
  assay(vsd_Object),
  batch=vsd_Object$batch,
  design=model.matrix(~condition, colData(vsd_Object))
  )
plotPCA(vsd_Object)
```

This makes sense: PC1 mostly separates *scute* overexpression from the rest, and PC2 is dominated by 'amount' of *daughterless* function (*da^RNAi^* < *control* < *da* < *da:da*). So we save this:
```{r save-vst-bcorr-counts}
outpath <- file.path(getwd(), 'output')
if ( !dir.exists(outpath) ) dir.create( outpath )
saveRDS(vsd_Object, file.path(outpath, 'vst_pseudocounts_batchCorrected.RDS'))
```


### Sample correlations


To further visualise the similarity of the samples we use the transformed, normalized counts and perform hierarchical clustering (with hidden dendrograms):
```{r sample-correlations, fig.height=6}
# Compute the correlation values between samples
vsd_cor_Object <- cor(assay(vsd_Object)) 

# heatmap
main.title <- 'RNAseq sample correlations'
## get sorted clusters
sort_hclust <- function(x) as.hclust(dendsort(as.dendrogram(x)))
mat_cluster_cols <- hclust(dist(t(vsd_cor_Object)))
mat_cluster_cols <- sort_hclust(mat_cluster_cols)
mat_cluster_rows <- hclust(dist(vsd_cor_Object))
mat_cluster_rows <- sort_hclust(mat_cluster_rows)
## mark the batches
annot_batch <- data.frame(batch = ifelse(test = targets$Batch == 'a',
                                         yes = 'batch A',
                                         no = 'batch B'))
rownames(annot_batch) <- rownames(vsd_cor_Object)
## get minimum correlation value, rounded for the legend
bot <- ceiling(min(vsd_cor_Object)*100)/100
## plot
pheatmap(
  # data
    mat               = vsd_cor_Object,
    scale             = "none", # otherwise numbers are changed
    cellwidth         = 15,
    cellheight        = 15,
  # title
    main              = main.title,
    fontsize          = 14,
    annotation        = dplyr::select(exptDesign, condition),
  # rows
    cluster_rows      = mat_cluster_rows,
    treeheight_row    = 25, # default is 50
    show_rownames     = TRUE,
    labels_row        = rownames(exptDesign),
    fontsize_row      = 9,
    annotation_row    = annot_batch,
  # cols
    cluster_cols      = mat_cluster_cols,
    treeheight_col    = 25,
    show_colnames     = TRUE,
    labels_col        = rownames(exptDesign),
    fontsize_col      = 9,
    angle_col         = 45,
  # legends
    legend_breaks     = c(bot, 1),
  # tiles
    color             = scico(255, palette='bamako'),
    border_color      = 'grey80')
```

This is somewhat redundant with PCA but also makes sense, so we will save the `vsd_Object` for the descriptive visualisations.
```{r save-pseudocounts}
saveRDS(vsd_Object, file.path(outpath, 'vsd.RDS'))
```


## 2.3 Visualise dispersion estimates


This will normalise the data, correct for dispersion (variance between replicates) and set data up for a differential comparison of any 2 conditions:
```{r analysis-object}
analysisObject = DESeq(exptObject)
```

After fitting the data to a negative binomial model, some basic QC consists of plotting the dispersion estimates using the `plotDispEsts()` function. The dispersion estimates are used to model the raw counts, assuming (1) that dispersions will generally decrease with increasing mean, and (2) that they should more or less follow the fitted line.
```{r plot-dispersion}
plotDispEsts(analysisObject)
```

This seems reasonably in order, so we move on to saving the expression data for further visualisation and analysis, and Supplementary Data:
```{r save-counts}
rawCounts        <- as.data.frame(counts(analysisObject, normalized=FALSE))
normalisedCounts <- as.data.frame(counts(analysisObject, normalized=TRUE))
saveRDS(rawCounts, file.path(outpath, 'rawCounts.RDS'))
saveRDS(normalisedCounts, file.path(outpath, 'normalisedCounts.RDS'))
```


## 2.4 Differential gene expression


Loop over the experimental conditions and save `DESeq2::results` as RDS data:
```{r get-dge, warning=FALSE}
# to get a more informative naming for the samples:
targets$sampleIDs <- names(rawCounts)
# conditions to be tested
test_conditions <- unique( targets[targets$Condition != 'Control',]$Condition )
test_names <- paste0(rep('Control_vs_',length(test_conditions)),test_conditions)
tests <- as.list(rep(NA, length(test_names)))
names(tests) <- test_names 

for (condtn in test_conditions) {
  # get the Counts for those conditions
  deData <- as.data.frame(results(analysisObject,
                                  contrast=c("condition", condtn, 'Control'), # Reference goes last!
                                  pAdjustMethod="BH"))
  # add column of ID
  deData <- cbind(data.frame('ensemblGeneID'=rownames(deData)), deData)
  # sort by pval
  deData <- deData[order(deData$pvalue), ] 
  # add gene symbol column and reorder columns
  deData <- merge(deData, dlist, by=0)
  deData <- deData[,c(1,ncol(deData),2:(ncol(deData)-1))]
  # save for later
  saveRDS(deData, file=file.path(outpath, paste0('Control_vs_', condtn, ".RDS")))
  # save as Supplementary data for publication
  cols <- c('gene_symbol', 'ensemblGeneID', 'log2FoldChange', 'padj')
  rownames(deData) <- NULL
  tests[[paste0('Control_vs_', condtn)]] <- deData %>% dplyr::select(all_of(cols))
}
```
(Note that in the parameter `contrast`, `'Control'` is last - I cannot find formal justification of this, but even with the `relevel`ling of the `condition` column of the experimental design object to make 'Control' the reference level, this needs to go last, or the log~2~FC values will have the opposite sign.)


### Supplementary Table S3


Now we just need to write the adjusted p-values and the log~2~(fold changes). That, with the raw counts, should be enough for anyone to do their own 'light' analysis without having to get any files from GEO.
```{r table-s3}
# add `rawCounts` to `tests`
tests <- rlist::list.append(tests, `Raw counts per gene per sample`=rawCounts)
# Supplementary data for publication
write_xlsx(tests, path = file.path('output', 'Table S3.xlsx'))
```

Finally, to have the experimental designed captured in a flexible manner for later on:
```{r save-targets}
targets$condition_md <- plyr::mapvalues(
  targets$Condition,
  from=unique(targets$Condition),
  to=c('*da^RNAi^*', '*da*', '*wild-type*', '*da:da*', '*scute*')
  )
targets$condition_md <- factor(
  targets$condition_md,
  c('*wild-type*', '*da*', '*da:da*','*da^RNAi^*', '*scute*')
  )
saveRDS(targets, file.path(outpath, 'targets.RDS'))
```
